Efficient Computation of Extended Sources

### 2021-06-14 repro/papers/sseprecond/paper.tex
discuss lam transpose

To illustrate these points, I have re-computed the source gather
estimates shown in Figures \ref{fig:preddinvhshh0} and
\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
via equation \ref{eqn:lamident} for this configuration is evaluated in
the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
model). The effect of aperture limitation is clearly diminished: the
second figure in each series compares the full-aperture pressure source
gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
approximated image of the corresponding pressure gather (Figure
\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
the last figures show that the approximated source gathers accurately
predict the point-source pressure gather at the receiver datum
$z_r=1000$ m.

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamnear} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Since successful preconditioning requires only approximate inversion,
use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
still yield a working preconditioner, and the former can be transposed
to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
     z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
 \end{equation}
Thus the model space weight operator introduced in the last section is
replaced by
\[
  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
  (\tilde{\Lambda}^+_{z_s})^T)
\]
\[
  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
\]
\[
  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
\]
\begin{equation}
  \label{eqn:wcomp}
  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
\end{equation}
with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.

@@

@@

>>>>

To illustrate these points, I have re-computed the source gather
estimates shown in Figures \ref{fig:preddinvhshh0} and
\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
via equation \ref{eqn:lamident} for this configuration is evaluated in
the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
model). The effect of aperture limitation is clearly diminished: the
second figure in each series compares the full-aperture pressure source
gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
approximated image of the corresponding pressure gather (Figure
\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
the last figures show that the approximated source gathers accurately
predict the point-source pressure gather at the receiver datum
$z_r=1000$ m.
As mentioned in the last section, computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamnear} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Of course, all of the examples so far show images of
$\tilde{\Lambda}^+_{z_s}$.

Since successful preconditioning requires only approximate inversion,
use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
still yield a working preconditioner, and the former can be transposed
to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
\end{equation}

Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
in Figure \ref{fig:dsrcphh0} under $(\tilde{\Lambda}^+_{z_s})^T$,
using the ``near'' traces at $z=2900$, that is, $\Delta z = 100$ m in
expression \ref{eqn:lamtransp}, and propagation in the
homogeneous model. Note
the close resemblance to the image of the same pressure gather under
$\tilde{\Lambda}^+_{z_s}$ displayed in Figure
\ref{fig:preddnshshh0}. The difference of these two images is
displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
images themselves. Since the propagation takes place entirely in a
region where all of the mechanical parameters are homogenous, I do not
offer a similar comparison for the lens model.

The model space weight operator $W_m$ introduced in the last section is
replaced by its asymptotic approximation
\[
  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
  (\tilde{\Lambda}^+_{z_s})^T)
\]
\[
  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
\]
\[
  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
\]
\begin{equation}
  \label{eqn:wcomp}
  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
\end{equation}
with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.

@@

Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
approximate source-to-pressure operator per equation \ref{eqn:wcomp},
applied once again to the pressure data in Figure
\ref{fig:dsrcp}. Note the resemblance to Figures
\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
asymptotic approximations of each other.

@@

\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
  ``near'' receiver traces at $z=2900$ m. Compare Figure
  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
  receivers are close, little aperture is lost in this case.}

\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
  array to implement formulat \ref{eqn:lamident}, and (b) source gather
  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
  Homogeneous model used in all
  propagations. Same color scale as in Figure \ref{fig:preddnshshh0}. }

\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
  under {\em transpose} of pressure-to-source operator
  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
  asymptotically symmetric, so the resemblance is not a surprise.}

\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
  on the same scale as these figures, showing that the asymptotic
  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
  length, time and frequency scales of these examples.}

\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
    under {\em symmetrized} pressure-to-source operator
    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
    model, using ``near'' receiver traces at $z=2900$ m. Compare
    Figure \ref{fig:preddnshshh0}.}
